---
title: "Causal Inference for linguistics"
author: "Stefano Coretta"
format: revealjs
knitr:
  opts_chunk:
    dev: png
    dev.args:
      bg: transparent
---

##  {background-image="../img/courtney-cook-66IZaW9LIpI-unsplash.jpg"}

[**"Correlation is not causation"**]{style="font-size: 60px"}

<br><br><br><br><br><br><br><br><br><br>

[Photo by <a href="https://unsplash.com/@courtneymcook?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText">Courtney Cook</a> on <a href="https://unsplash.com/photos/four-round-white-and-brown-ornaments-66IZaW9LIpI?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText">Unsplash</a>]{style="font-size: 8px;"}

## Clothing and ice cream

<br><br><br><br>

::: {style="text-align: center; font-size: 50px;"}
🎽 shorter clothing ↔️ more ice cream 🍨
:::

## Counfound: temperature

<br><br><br>

::: {style="text-align: center; font-size: 50px;"}
🌡️ higher temperature ➡️ shorter clothing 🎽

<br>

🌡️ higher temperature ➡️ more ice cream 🍨
:::

## Causal effects

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(ggdag)
library(dagitty)
```

```{r}
#| label: dag-temp

temp <- dagify(
  C ~ T,
  I ~ T,
  coords = list(
    x = c(C = 1, I = 2, T = 1.5),
    y = c(C = 1, I = 1, T = 2)
  )
)

ggplot(temp, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point() +
  geom_dag_text() +
  theme_dag(
    base_size = 20,
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )
```

## Directed acyclic graphs (DAGs)

::: {.callout-note appearance="simple"}
-   **Graphical causal theory**.

-   **Nodes** represent variables.

-   **Edges** (arrows) represent causal effects (*directed*).

-   Causality flows "linearly", no circular causality (*acylic*).
:::

## Sentence length and subordination frequency

```{r}
#| label: dag-pandermaat

temp <- dagify(
  L ~ C,
  S ~ C,
  coords = list(
    x = c(L = 1, S = 2, C = 1.5),
    y = c(L = 1, S = 1, C = 2)
  ),
  exposure = "L",
  outcome = "S"
)

ggplot(temp, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point() +
  geom_dag_text() +
  theme_dag(
    base_size = 20,
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )

```
