---
title: "Causal Inference for linguistics"
author: "Stefano Coretta"
format: revealjs
knitr:
  opts_chunk:
    dev: png
    dev.args:
      bg: transparent
bibliography: references.bib
---

##  {background-image="../img/courtney-cook-66IZaW9LIpI-unsplash.jpg"}

[**"Correlation is not causation"**]{style="font-size: 60px"}


::: aside
[Photo by <a href="https://unsplash.com/@courtneymcook?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText">Courtney Cook</a> on <a href="https://unsplash.com/photos/four-round-white-and-brown-ornaments-66IZaW9LIpI?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText">Unsplash</a>]{style="font-size: 8px;"}
:::

## Clothing and ice cream

<br><br><br><br>

::: {style="text-align: center; font-size: 50px;"}
🎽 shorter clothing ↔️ more ice cream 🍨
:::

## Counfound: temperature

<br><br><br>

::: {style="text-align: center; font-size: 50px;"}
🌡️ higher temperature ➡️ shorter clothing 🎽

<br>

🌡️ higher temperature ➡️ more ice cream 🍨
:::

## Causal effects

```{r}
#| label: setup
#| include: false

library(tidyverse)
theme_set(theme_light())
library(ggdag)
library(dagitty)
library(patchwork)

gg_dag <- function(dag, base_size = 20) {
  ggplot(dag, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point() +
  geom_dag_text() +
  theme_dag(
    base_size = base_size,
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )
  # expand_plot(expand_x = c(0.1, 0.1), expand_y = c(0.1, 0.1))
}
```

```{r}
#| label: dag-temp

temp <- dagify(
  C ~ T,
  I ~ T,
  coords = list(
    x = c(C = 1, I = 2, T = 1.5),
    y = c(C = 1, I = 1, T = 2)
  )
)

gg_dag(temp)
```

## Directed acyclic graphs (DAGs)

::: {.callout-note appearance="simple"}
-   **Graphical causal theory**.

-   **Nodes** represent variables.

-   **Edges** (arrows) represent causal effects (*directed*).

-   Causality flows "linearly", no circular causality (*acylic*).
:::

```{r}
#| label: dag-temp-2
#| fig-align: center

gg_dag(temp)
```

## Sentence length and subordination frequency

::: fragment

```{r}
#| label: dag-pandermaat
#| fig-align: center

temp <- dagify(
  L ~ C,
  S ~ C,
  coords = list(
    x = c(L = 1, S = 2, C = 1.5),
    y = c(L = 1, S = 1, C = 2)
  ),
  exposure = "L",
  outcome = "S"
)

gg_dag(temp)

```

:::

::: aside
C = discourse complexity, L = sentence length, S = subordination frequency. [@pandermaat2017]{.small-ref}
:::

## Four path types

```{r}
#| label: elemental
#| layout-ncol: 2
#| layout-nrow: 2

# Adapted from https://solomon.quarto.pub/sr2/06.html#shutting-the-backdoor

d1 <- dagify(
  X ~ Z,
  Y ~ Z,
  coords = tibble(name = c("X", "Y", "Z"),
                  x = c(1, 3, 2),
                  y = c(1, 1, 2)))

d2 <- dagify(
  Z ~ X,
  Y ~ Z,
  coords = tibble(name = c("X", "Y", "Z"),
                  x = c(1, 3, 2),
                  y = c(2, 1, 1.5)))

d3 <- dagify(
  Z ~ X + Y,
  coords = tibble(name = c("X", "Y", "Z"),
                  x = c(1, 3, 2),
                  y = c(2, 2, 1)))

d4 <- dagify(
  Z ~ X + Y,
  D ~ Z,
  coords = tibble(name = c("X", "Y", "Z", "D"),
                  x = c(1, 3, 2, 2),
                  y = c(2, 2, 2, 1.05)))

p1 <- gg_dag(d1) + labs(subtitle = "The Fork")
p2 <- gg_dag(d2) + labs(subtitle = "The Pipe")
p3 <- gg_dag(d3) + labs(subtitle = "The Collider")
p4 <- gg_dag(d4) + labs(subtitle = "The Descendant")

# (p1 | p2 | p3 | p4) &
#   theme(plot.subtitle = element_text(hjust = 0.5)) &
#   plot_annotation(title = "The four elemental confounds") 

p1
p2
p3
p4
```

## The Fork: Z is a confound

```{r}
#| label: elemental-fork

gg_dag(d1)
```

## Confound: simulate

::: .columns

::: {.column width="40%"}

```{r}
#| label: confound-sim
#| echo: true

n <- 100

z <- rnorm(100)

x <- 0.4 * z + rnorm(100, 0, 0.1)
y <- 0.7 * z + rnorm(100, 0, 0.1)
```

:::

::: {.column width="60%" .fragment}

```{r}
#| label: confound-plot
#| fig-asp: 1
#| fig-align: center

tibble(x, y) |> 
  ggplot(aes(x, y)) +
  geom_point() +
  theme(text = element_text(size = 20))
```


:::

:::

## Confound: example



## References
