---
title: "Causal Inference for linguistics"
author: "Stefano Coretta"
format: revealjs
bibliography: references.bib
---

##  {background-image="../img/courtney-cook-66IZaW9LIpI-unsplash.jpg"}

[**"Correlation is not causation"**]{style="font-size: 60px"}


::: aside
[Photo by <a href="https://unsplash.com/@courtneymcook?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText">Courtney Cook</a> on <a href="https://unsplash.com/photos/four-round-white-and-brown-ornaments-66IZaW9LIpI?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText">Unsplash</a>]{style="font-size: 8px;"}
:::

## Clothing and ice cream

<br><br><br><br>

::: {style="text-align: center; font-size: 50px;"}
🎽 shorter clothing ↔️ more ice cream 🍨
:::

## Counfound: temperature

<br><br><br>

::: {style="text-align: center; font-size: 50px;"}
🌡️ higher temperature ➡️ shorter clothing 🎽

<br>

🌡️ higher temperature ➡️ more ice cream 🍨
:::

## Causal effects

```{r}
#| label: setup
#| include: false
#| fig-align: center

library(tidyverse)
theme_set(theme_light())
library(ggdag)
library(dagitty)
library(patchwork)
library(GGally)

gg_dag <- function(dag, base_size = 20) {
  ggplot(dag, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point() +
  geom_dag_text() +
  theme_dag(
    base_size = base_size,
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )
  # expand_plot(expand_x = c(0.1, 0.1), expand_y = c(0.1, 0.1))
}
```

```{r}
#| label: dag-temp

temp <- dagify(
  C ~ T,
  I ~ T,
  coords = list(
    x = c(C = 1, I = 2, T = 1.5),
    y = c(C = 1, I = 1, T = 2)
  )
)

gg_dag(temp)
```

## Directed acyclic graphs (DAGs)

::: {.callout-note appearance="simple"}
-   **Graphical causal theory**.

-   **Nodes** represent variables.

-   **Edges** (arrows) represent causal effects (*directed*).

-   Causality flows "linearly", no circular causality (*acylic*).
:::

```{r}
#| label: dag-temp-2

gg_dag(temp)
```

## Sentence length and subordination frequency

::: fragment

```{r}
#| label: dag-pandermaat

temp <- dagify(
  L ~ C,
  S ~ C,
  coords = list(
    x = c(L = 1, S = 2, C = 1.5),
    y = c(L = 1, S = 1, C = 2)
  ),
  exposure = "L",
  outcome = "S"
)

gg_dag(temp)

```

:::

::: aside
C = discourse complexity, L = sentence length, S = subordination frequency. [@pandermaat2017]{.small-ref}
:::

## Four path types

```{r}
#| label: elemental
#| layout-ncol: 2
#| layout-nrow: 2

# Adapted from https://solomon.quarto.pub/sr2/06.html#shutting-the-backdoor

d1 <- dagify(
  X ~ Z,
  Y ~ Z,
  coords = tibble(name = c("X", "Y", "Z"),
                  x = c(1, 3, 2),
                  y = c(1, 1, 2)))

d2 <- dagify(
  Z ~ X,
  Y ~ Z,
  coords = tibble(name = c("X", "Y", "Z"),
                  x = c(1, 3, 2),
                  y = c(2, 1, 1.5)))

d3 <- dagify(
  Z ~ X + Y,
  coords = tibble(name = c("X", "Y", "Z"),
                  x = c(1, 3, 2),
                  y = c(2, 2, 1)))

d4 <- dagify(
  Z ~ X + Y,
  D ~ Z,
  coords = tibble(name = c("X", "Y", "Z", "D"),
                  x = c(1, 3, 2, 2),
                  y = c(2, 2, 2, 1.05)))

p1 <- gg_dag(d1) + labs(subtitle = "The Fork")
p2 <- gg_dag(d2) + labs(subtitle = "The Pipe")
p3 <- gg_dag(d3) + labs(subtitle = "The Collider")
p4 <- gg_dag(d4) + labs(subtitle = "The Descendant")

# (p1 | p2 | p3 | p4) &
#   theme(plot.subtitle = element_text(hjust = 0.5)) &
#   plot_annotation(title = "The four elemental confounds") 

p1
p2
p3
p4
```

## The Fork: Z is a confound

```{r}
#| label: elemental-fork

gg_dag(d1)
```

## Confound: simulate

::: .columns

::: {.column width="40%"}

```{r}
#| label: confound-sim
#| echo: true

set.seed(9988)

n <- 100

z <- rnorm(100)

x <- 0.4 * z + rnorm(100, 0, 0.1)
y <- 0.7 * z + rnorm(100, 0, 0.1)
```

:::

::: {.column width="60%" .fragment}

```{r}
#| label: confound-plot
#| fig-asp: 1

conf <- tibble(x, z, y)

conf |> 
  ggplot(aes(x, y)) +
  geom_point() +
  theme(text = element_text(size = 20))
```


:::

:::



## Confound: greater proficiency and code-switching

```{r}
#| label: pro-switch

pro_switch <- dagify(
  S ~ P,
  coords = tibble(
    name = c("P", "S"),
    x = c(1, 2),
    y = c(1, 1)
  )
)

gg_dag(pro_switch)
```

::: aside
P = proficiency, S = code-switching. [@schachingertenes2023]{.small-ref}
:::

## Confound: multicultural context

```{r}
#| label: pro-switch-confound

pro_switch <- dagify(
  P ~ C,
  S ~ C,
  coords = tibble(
    name = c("P", "S", "C"),
    x = c(1, 3, 2),
    y = c(1, 1, 2)
  )
)

gg_dag(pro_switch)

```

::: aside
C = multicultural context, P = proficiency, S = code-switching. [@schachingertenes2023]{.small-ref}
:::

## The Pipe: Z is a moderator

```{r}
#| label: elemental-pipe

gg_dag(d2)
```

## Moderator: simulate

::: .columns

::: {.column width="40%"}

```{r}
#| label: moderator-sim
#| echo: true

set.seed(9988)

n <- 100

x <- rnorm(100)

z <- 0.4 * x + rnorm(100, 0, 0.1)
y <- 0.7 * z + rnorm(100, 0, 0.1)
```

:::

::: {.column width="60%" .fragment}

```{r}
#| label: moderator-plot
#| fig-asp: 1

mod <- tibble(x, z, y)

mod |> 
  ggplot(aes(x, y)) +
  geom_point() +
  theme(text = element_text(size = 20))
```


:::

:::

## Moderator: bilingualism and executive function

```{r}
#| label: bi-exec
#| fig-asp: 1
#| fig-width: 5

bi_exec <- dagify(
  E ~ B,
  coords = tibble(
    name = c("B", "E"),
    x = c(1, 2),
    y = c(2, 1)
  )
)

gg_dag(bi_exec)
```

::: aside
B = bilingualism, E = executive function. [@paap2014]{.small-ref}
:::

## Moderator: code-switching

```{r}
#| label: bi-exec-mod
#| fig-asp: 1
#| fig-width: 5

bi_exec_mod <- dagify(
  E ~ S, S ~ B,
  coords = tibble(
    name = c("B", "E", "S"),
    x = c(1, 2, 2),
    y = c(2, 1, 2)
  )
)

gg_dag(bi_exec_mod)
```

::: aside
B = bilingualism, S = code-switching, E = executive function. [@paap2014]{.small-ref}
:::

## Confound vs moderator

```{r}
#| label: conf-mod
#| fig-asp: 1
#| layout-ncol: 2

ggpairs(conf, upper = list(continuous = wrap(ggally_cor, stars = FALSE))) +
  labs(title = "Confound") +
  theme(
    text = element_text(size = 20),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )

ggpairs(mod, upper = list(continuous = wrap(ggally_cor, stars = FALSE))) +
  labs(title = "Moderator") +
  theme(
    text = element_text(size = 20),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )
```

## The Collider: Z is collider

```{r}
#| label: elemental-coll

gg_dag(d3)
```

## Collider: newsworthiness and trustworthiness

```{r}
#| label: new-trust

new_trust <- dagify(
  S ~ N + `T`,
  coords = tibble(name = c("N", "T", "S"),
                  x = c(1, 3, 2),
                  y = c(2, 2, 1)))

gg_dag(new_trust)
```


::: aside
N = newsworthiness, T = trustworthiness, S = selected for funding. [@mcelreath2020]{.small-ref}
:::

## Collider: simulate

```{r}
#| label: collider-sim
#| echo: true

# Adapted from https://solomon.quarto.pub/sr2/06.html#overthinking-simulated-science-distortion.
# Example from McElreath *Statistical Rethinking* (2nd ed).

set.seed(1914)

n <- 200  # Number of grant proposals
p <- 0.1  # Proportion to select

# Uncorrelated newsworthiness and trustworthiness
d <- tibble(
  newsworthiness = rnorm(n, mean = 0, sd = 1),
  trustworthiness = rnorm(n, mean = 0, sd = 1)
) |> 
  # total_score
  mutate(total_score = newsworthiness + trustworthiness) |> 
  # Select top 10% of combined scores
  mutate(selected = ifelse(total_score >= quantile(total_score, 1 - p), TRUE, FALSE))
```

## Collider: simulate (plot)

```{r}
#| label: coolider-plot
#| message: false
#| fig-asp: 1

# Adapted from https://solomon.quarto.pub/sr2/06.html#overthinking-simulated-science-distortion.

# We'll need this for the annotation
text <- tibble(
  newsworthiness  = c(2, 1), 
  trustworthiness = c(2, -2.25),
  selected        = c(TRUE, FALSE),
  label           = c("selected", "rejected"))

d |> 
  ggplot(aes(x = newsworthiness, y = trustworthiness, color = selected)) +
  geom_point(aes(shape = selected), alpha = 3/4, size = 3) +
  geom_label(data = text, aes(label = label), size = 20 / .pt) +
  geom_smooth(data = d |> filter(selected == TRUE),
              method = "lm", fullrange = T, se = F,
              color = "darkblue", linewidth = 1/2) +
  scale_x_continuous(limits = c(-3, 3.9), expand = c(0, 0)) +
  scale_color_manual(values = c("black", "darkblue")) +
  scale_shape_manual(values = c(1, 19)) +
  coord_cartesian(ylim = range(d$trustworthiness)) +
  theme(legend.position = "none", text = element_text(size = 20))
```


## Confound: regression

::: columns
::: {.column}
```{r}
#| label: conf-lm-1
#| echo: true

conf_lm_1 <- lm(y ~ x, data = conf)
```

::: fragment
<br>

```{r}
coef(summary(conf_lm_1))[, 1:2] |> round(3)
```
:::

:::

::: {.column .fragment}
```{r}
#| label: conf-lm-2
#| echo: true

conf_lm_2 <- lm(y ~ x + z, data = conf)
```

::: fragment
<br>

```{r}
coef(summary(conf_lm_2))[, 1:2] |> round(3)
```
:::

:::
:::

## Moderator: regression

::: columns
::: {.column}
```{r}
#| label: mod-lm-1
#| echo: true

mod_lm_1 <- lm(y ~ x, data = mod)
```

::: fragment
<br>

```{r}
coef(summary(mod_lm_1))[, 1:2] |> round(3)
```
:::

:::

::: {.column .fragment}
```{r}
#| label: mod-lm-2
#| echo: true

mod_lm_2 <- lm(y ~ x + z, data = mod)
```

::: fragment
<br>

```{r}
coef(summary(mod_lm_2))[, 1:2] |> round(3)
```
:::

:::
:::




## References
