---
title: "Causal Inference"
subtitle: "Correlation *is* causation"
author: "Stefano Coretta"
date: 20/05/2026
institute: University of Edinburgh
format: revealjs
bibliography: references.bib
---

##  {background-image="../img/courtney-cook-66IZaW9LIpI-unsplash.jpg"}

[**"Correlation is not causation"**]{style="font-size: 60px"}

::: aside
[Photo by <a href="https://unsplash.com/@courtneymcook?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText">Courtney Cook</a> on <a href="https://unsplash.com/photos/four-round-white-and-brown-ornaments-66IZaW9LIpI?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText">Unsplash</a>]{style="font-size: 8px;"}
:::

## Clothing and ice cream

<br><br><br><br>

::: {style="text-align: center; font-size: 50px;"}
🎽 shorter clothing ↔️ more ice cream 🍨
:::

## Confound: temperature

<br><br><br>

::: {style="text-align: center; font-size: 50px;"}
🌡️ higher temperature ➡️ shorter clothing 🎽

<br>

🌡️ higher temperature ➡️ more ice cream 🍨
:::

## Causal effects

```{r}
#| label: setup
#| include: false
#| fig-align: center

library(tidyverse)
theme_set(theme_light())
library(ggdag)
library(dagitty)
library(patchwork)
library(GGally)

gg_dag <- function(dag) {
  ggplot(dag, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point() +
  geom_dag_text() +
  theme_dag(
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )
  # expand_plot(expand_x = c(0.1, 0.1), expand_y = c(0.1, 0.1))
}
```

```{r}
#| label: dag-temp

temp <- dagify(
  C ~ T,
  I ~ T,
  coords = list(
    x = c(C = 1, I = 2, T = 1.5),
    y = c(C = 1, I = 1, T = 2)
  )
)

gg_dag(temp)
```

## Directed acyclic graphs (DAGs)

::: {.callout-note appearance="simple"}
- **Graphical causal theory**.

- **Nodes** represent variables.

- **Edges** (arrows) represent causal effects (*directed*).

- Causality flows "linearly", no circular causality (*acylic*).
:::

```{r}
#| label: dag-temp-2

gg_dag(temp)
```

## Sentence length and subordination frequency

::: fragment
```{r}
#| label: dag-pandermaat

temp <- dagify(
  L ~ C,
  S ~ C,
  coords = list(
    x = c(L = 1, S = 2, C = 1.5),
    y = c(L = 1, S = 1, C = 2)
  ),
  exposure = "L",
  outcome = "S"
)

gg_dag(temp)

```
:::

::: aside
C = discourse complexity, L = sentence length, S = subordination frequency.
@pandermaat2017
:::

## Four path types

```{r}
#| label: elemental
#| layout-ncol: 2
#| layout-nrow: 2

# Adapted from https://solomon.quarto.pub/sr2/06.html#shutting-the-backdoor

d1 <- dagify(
  X ~ Z,
  Y ~ Z,
  coords = tibble(name = c("X", "Y", "Z"),
                  x = c(1, 3, 2),
                  y = c(1, 1, 2)))

d2 <- dagify(
  Z ~ X,
  Y ~ Z,
  coords = tibble(name = c("X", "Y", "Z"),
                  x = c(1, 3, 2),
                  y = c(2, 1, 1.5)))

d3 <- dagify(
  Z ~ X + Y,
  coords = tibble(name = c("X", "Y", "Z"),
                  x = c(1, 3, 2),
                  y = c(2, 2, 1)))

d4 <- dagify(
  Z ~ X + Y,
  D ~ Z,
  coords = tibble(name = c("X", "Y", "Z", "D"),
                  x = c(1, 3, 2, 2),
                  y = c(2, 2, 2, 1.05)))

p1 <- gg_dag(d1) + labs(subtitle = "The Fork")
p2 <- gg_dag(d2) + labs(subtitle = "The Pipe")
p3 <- gg_dag(d3) + labs(subtitle = "The Collider")
p4 <- gg_dag(d4) + labs(subtitle = "The Descendant")

# (p1 | p2 | p3 | p4) &
#   theme(plot.subtitle = element_text(hjust = 0.5)) &
#   plot_annotation(title = "The four elemental confounds") 

p1
p2
p3
p4
```

## The Fork: Z is a confound

```{r}
#| label: elemental-fork

gg_dag(d1)
```

## Confound: simulate

::::: columns
::: {.column width="40%"}
```{r}
#| label: confound-sim
#| echo: true

set.seed(9988)

n <- 100

z <- rnorm(100)

x <- 0.4 * z + rnorm(100, 0, 0.1)
y <- 0.7 * z + rnorm(100, 0, 0.1)
```
:::

::: {.column .fragment width="60%"}
```{r}
#| label: confound-plot
#| fig-asp: 1

conf <- tibble(x, z, y)

conf |> 
  ggplot(aes(x, y)) +
  geom_point() +
  theme(text = element_text(size = 20))
```
:::
:::::

## Confound: greater proficiency and code-switching

```{r}
#| label: pro-switch
#| fig-asp: 1
#| fig-width: 5

pro_switch <- dagify(
  S ~ P,
  coords = tibble(
    name = c("P", "S"),
    x = c(1, 2),
    y = c(1, 1)
  )
)

gg_dag(pro_switch) +
  geom_point(aes(x = 1.5, y = 2), colour = "transparent")
```

::: aside
P = proficiency, S = code-switching.
@schachingertenes2023
:::

## Confound: multicultural context

```{r}
#| label: pro-switch-confound
#| fig-asp: 1
#| fig-width: 5

pro_switch <- dagify(
  P ~ C,
  S ~ C,
  coords = tibble(
    name = c("P", "S", "C"),
    x = c(1, 3, 2),
    y = c(1, 1, 2)
  )
)

gg_dag(pro_switch)

```

::: aside
C = multicultural context, P = proficiency, S = code-switching.
@schachingertenes2023
:::

## The Pipe: Z is a mediator

```{r}
#| label: elemental-pipe

gg_dag(d2)
```

## Mediator: simulate

::::: columns
::: {.column width="40%"}
```{r}
#| label: mediator-sim
#| echo: true

set.seed(9988)

n <- 100

x <- rnorm(100)

z <- 0.4 * x + rnorm(100, 0, 0.1)
y <- 0.7 * z + rnorm(100, 0, 0.1)
```
:::

::: {.column .fragment width="60%"}
```{r}
#| label: mediator-plot
#| fig-asp: 1

mod <- tibble(x, z, y)

mod |> 
  ggplot(aes(x, y)) +
  geom_point() +
  theme(text = element_text(size = 20))
```
:::
:::::

## Mediator: bilingualism and executive function

```{r}
#| label: bi-exec
#| fig-asp: 1
#| fig-width: 5

bi_exec <- dagify(
  E ~ B,
  coords = tibble(
    name = c("B", "E"),
    x = c(1, 2),
    y = c(2, 1)
  )
)

gg_dag(bi_exec)
```

::: aside
B = bilingualism, E = executive function.
@paap2014
:::

## Mediator: code-switching

```{r}
#| label: bi-exec-mod
#| fig-asp: 1
#| fig-width: 5

bi_exec_mod <- dagify(
  E ~ S, S ~ B,
  coords = tibble(
    name = c("B", "E", "S"),
    x = c(1, 2, 2),
    y = c(2, 1, 2)
  )
)

gg_dag(bi_exec_mod)
```

::: aside
B = bilingualism, S = code-switching, E = executive function.
@paap2014
:::

## Confound vs Mediator

```{r}
#| label: conf-mod
#| fig-asp: 1
#| layout-ncol: 2

ggpairs(conf, upper = list(continuous = wrap(ggally_cor, stars = FALSE))) +
  labs(title = "Confound") +
  theme(
    text = element_text(size = 20),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )

ggpairs(mod, upper = list(continuous = wrap(ggally_cor, stars = FALSE))) +
  labs(title = "Mediator") +
  theme(
    text = element_text(size = 20),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )
```

## The Collider: Z is collider

```{r}
#| label: elemental-coll

gg_dag(d3)
```

## Collider: newsworthiness and trustworthiness

```{r}
#| label: new-trust

new_trust <- dagify(
  S ~ N + `T`,
  coords = tibble(name = c("N", "T", "S"),
                  x = c(1, 3, 2),
                  y = c(2, 2, 1)))

gg_dag(new_trust)
```

::: aside
N = newsworthiness, T = trustworthiness, S = selected for funding.
@mcelreath2020
:::

## Collider: simulate

```{r}
#| label: collider-sim
#| echo: true

# Adapted from https://solomon.quarto.pub/sr2/06.html#overthinking-simulated-science-distortion.
# Example from McElreath *Statistical Rethinking* (2nd ed).

set.seed(1914)

n <- 200  # Number of grant proposals
p <- 0.1  # Proportion to select

# Uncorrelated newsworthiness and trustworthiness
coll <- tibble(
  newsworthiness = rnorm(n, mean = 0, sd = 1),
  trustworthiness = rnorm(n, mean = 0, sd = 1)
) |> 
  # total_score
  mutate(total_score = newsworthiness + trustworthiness) |> 
  # Select top 10% of combined scores
  mutate(selected = ifelse(total_score >= quantile(total_score, 1 - p), TRUE, FALSE))
```

## Collider: simulate (plot)

```{r}
#| label: collider-plot
#| message: false
#| fig-asp: 1

# Adapted from https://solomon.quarto.pub/sr2/06.html#overthinking-simulated-science-distortion.

# We'll need this for the annotation
text <- tibble(
  newsworthiness  = c(2, 1), 
  trustworthiness = c(2, -2.25),
  selected        = c(TRUE, FALSE),
  label           = c("selected", "rejected"))

coll |> 
  ggplot(aes(x = newsworthiness, y = trustworthiness, color = selected)) +
  geom_point(aes(shape = selected), alpha = 3/4, size = 3) +
  geom_label(data = text, aes(label = label), size = 20 / .pt) +
  geom_smooth(data = coll |> filter(selected == TRUE),
              method = "lm", fullrange = T, se = F,
              color = "darkblue", linewidth = 1/2) +
  scale_x_continuous(limits = c(-3, 3.9), expand = c(0, 0)) +
  scale_color_manual(values = c("black", "darkblue")) +
  scale_shape_manual(values = c(1, 19)) +
  coord_cartesian(ylim = range(coll$trustworthiness)) +
  theme(legend.position = "none", text = element_text(size = 20))
```

## Collider: verb frequency and irregularity

```{r}
#| label: freq-irr
#| fig-asp: 1
#| fig-width: 5

freq_irr <- dagify(
  I ~ `F`,
  coords = tibble(
    name = c("F", "I"),
    x = c(1, 2),
    y = c(1, 1)
  )
)

gg_dag(freq_irr) +
  geom_point(aes(x = 1.5, y = -1), color = "transparent")
```

::: aside
F = frequency, I = irregularity.
@wu2019
:::

## Collider: verb saliency

```{r}
#| label: freq-irr-coll
#| fig-asp: 1
#| fig-width: 5

bi_exec_mod <- dagify(
  S ~ `F`, S ~ I,
  coords = tibble(
    name = c("F", "I", "S"),
    x = c(1, 3, 2),
    y = c(2, 2, 1)
  )
)

gg_dag(bi_exec_mod)
```

::: aside
F = frequency, S = saliency, I = irregularity.
@wu2019
:::

##  {background-image="../img/patrick-robert-doyle-EkBCBo7vmM4-unsplash.jpg"}

[Regression for causation]{.absolute .r-fit-text top="450" left="0" style="color: white;"}

::: aside
[Photo by [Patrick Robert Doyle](https://unsplash.com/@teapowered?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText) on [Unsplash](https://unsplash.com/photos/white-and-black-mountain-EkBCBo7vmM4?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText)]{style="font-size: 8px;"}
:::

## Colonial history (confound)

```{r}
#| label: colonial-dag
#| fig-asp: 1
#| fig-width: 5

colonial <- dagify(
  E ~ C, L ~ C,
  coords = tibble(
    name = c("C", "E", "L"),
    x = c(1.5, 1, 2),
    y = c(2, 1, 1)
  )
)

colonial_dag <- ggplot(colonial, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(
    end_cap = ggraph::circle(15, "mm"),
    start_cap = ggraph::circle(15, "mm")
  ) +
  geom_dag_point(size = 20) +
  geom_dag_text(size = 20 / .pt) +
  expand_plot(expand_y = c(0.3, 0.3)) +
  theme_dag(
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )

colonial_dag
```

::: aside
C = colonial history, E = economic development, L = language endangerment.
:::

## Colonial history: simulate {.nostretch}

```{r}
#| label: colonial-dag-2
#| fig-asp: 0.25
#| out-width: 40%

colonial_dag
```

```{r}
#| label: colonial-sim
#| echo: true

set.seed(9988)

n <- 100

C <- rnorm(100)

E <- 0.4 * C + rnorm(100, 0, 0.1)
L <- 0.7 * C + rnorm(100, 0, 0.1)

col_df <- tibble(C, E, L)
```

::: aside
C = colonial history, E = economic development, L = language endangerment.
:::

## Colonial history: regression {.nostretch}

```{r}
#| label: colonial-dag-3
#| fig-asp: 0.25
#| out-width: 40%

colonial_dag
```

::::::: columns
:::: column
```{r}
#| label: conf-lm-1
#| echo: true

conf_lm_1 <- lm(L ~ E, data = col_df)
```

::: {.fragment style="font-size: 30pt;"}
<br>

```{r}
coef(summary(conf_lm_1))[, 1:2] |> round(3)
```
:::
::::

:::: {.column .fragment}
```{r}
#| label: conf-lm-2
#| echo: true

conf_lm_2 <- lm(L ~ E + C, data = col_df)
```

::: {.fragment style="font-size: 30pt;"}
<br>

```{r}
coef(summary(conf_lm_2))[, 1:2] |> round(3)
```
:::
::::
:::::::

::: aside
C = colonial history, E = economic development, L = language endangerment.
:::

## Plant reliance (mediator)

```{r}
#| label: plant-dag
#| fig-asp: 1
#| fig-width: 5

plant <- dagify(
  R ~ B, P ~ R,
  coords = tibble(
    name = c("B", "R", "P"),
    x = c(1, 2, 3),
    y = c(3, 2, 1)
  )
)

plant_dag <- ggplot(plant, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(
    end_cap = ggraph::circle(15, "mm"),
    start_cap = ggraph::circle(15, "mm")
  ) +
  geom_dag_point(size = 20) +
  geom_dag_text(size = 20 / .pt) +
  expand_plot(expand_y = c(0.3, 0.3)) +
  theme_dag(
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )

plant_dag
```

::: aside
B = biodiversity, R = reliance on plants, P = number of plant names.
:::

## Plant reliance: simulate {.nostretch}

```{r}
#| label: plant-dag-2
#| fig-asp: 0.25
#| out-width: 40%

plant_dag
```

```{r}
#| label: plant-sim
#| echo: true

set.seed(9988)

n <- 100

B <- rnorm(100)

R <- 0.4 * B + rnorm(100, 0, 0.1)
P <- 0.7 * R + rnorm(100, 0, 0.1)

pla_df <- tibble(B, R, P)
```

## Plant reliance: regression {.nostretch}

```{r}
#| label: plant-dag-3
#| fig-asp: 0.25
#| out-width: 40%

plant_dag
```

::::::: columns
:::: column
```{r}
#| label: mod-lm-1
#| echo: true

mod_lm_1 <- lm(P ~ B, data = pla_df)
```

::: {.fragment style="font-size: 30pt;"}
<br>

```{r}
coef(summary(mod_lm_1))[, 1:2] |> round(3)
```
:::
::::

:::: {.column .fragment}
```{r}
#| label: mod-lm-2
#| echo: true

mod_lm_2 <- lm(P ~ B + R, data = pla_df)
```

::: {.fragment style="font-size: 30pt;"}
<br>

```{r}
coef(summary(mod_lm_2))[, 1:2] |> round(3)
```
:::
::::
:::::::

## Regression: collider {.nostretch}

```{r}
#| label: collider-dag
#| fig-asp: 0.25
#| out-width: 40%

ggplot(d3, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(
    end_cap = ggraph::circle(15, "mm"),
    start_cap = ggraph::circle(15, "mm")
  ) +
  geom_dag_point(size = 20) +
  geom_dag_text(size = 20 / .pt) +
  expand_plot(expand_y = c(0.3, 0.3)) +
  theme_dag(
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )
```

::::::: {.columns .fragment}
:::: column
```{r}
#| label: coll-lm-1
#| echo: true

coll_lm_1 <- lm(
  trustworthiness ~ newsworthiness,
  data = coll
)
```

::: {.fragment style="font-size: 30pt;"}
<br>

```{r}
coef(summary(coll_lm_1))[, 1:2] |> round(3)
```
:::
::::

:::: {.column .fragment}
```{r}
#| label: coll-lm-2
#| echo: true

coll_lm_2 <- lm(
  trustworthiness ~ newsworthiness + selected,
  data = coll
)
```

::: {.fragment style="font-size: 30pt;"}
<br>

```{r}
coef(summary(coll_lm_2))[, 1:2] |> round(3)
```
:::
::::
:::::::

## Conditional independences {.nostretch}

```{r}
#| label: pro-switch-1
#| fig-asp: 0.5
#| out-width: 40%

pro_switch_1 <- dagify(
  P ~ C,
  S ~ C + P,
  coords = tibble(
    name = c("P", "S", "C"),
    x = c(2, 1.5, 1),
    y = c(2, 1, 2)
  )
)

ps1 <- ggplot(pro_switch_1, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(
    end_cap = ggraph::circle(15, "mm"),
    start_cap = ggraph::circle(15, "mm")
  ) +
  geom_dag_point(size = 20) +
  geom_dag_text(size = 20 / .pt) +
  expand_plot(expand_y = c(0.1, 0.1)) +
  theme_dag(
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )

ps1
```

. . .

::::: columns
::: column
$P \not\!\perp\!\!\!\perp C$

$S \not\!\perp\!\!\!\perp C$

$S \not\!\perp\!\!\!\perp P$
:::

::: {.column .fragment}
$\perp\!\!\!\perp$ = "independent"

$\not\!\perp\!\!\!\perp$ = "not independent"
:::
:::::

::: aside
C = multicultural context, P = proficiency, S = code-switching.
@schachingertenes2023
:::

## Conditional independences {.nostretch}

```{r}
#| label: pro-switch-2
#| fig-asp: 0.5
#| out-width: 40%

pro_switch_2 <- dagify(
  P ~ C,
  S ~ C,
  coords = tibble(
    name = c("P", "S", "C"),
    x = c(2, 1.5, 1),
    y = c(2, 1, 2)
  )
)

ps2 <- ggplot(pro_switch_2, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(
    end_cap = ggraph::circle(15, "mm"),
    start_cap = ggraph::circle(15, "mm")
  ) +
  geom_dag_point(size = 20) +
  geom_dag_text(size = 20 / .pt) +
  expand_plot(expand_y = c(0.1, 0.1)) +
  theme_dag(
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )

ps2
```

. . .

::::: columns
::: {.column width="20%"}
$P \not\!\perp\!\!\!\perp C$

$S \not\!\perp\!\!\!\perp C$

$S \not\!\perp\!\!\!\perp P | C$
:::

::: {.column .fragment width="80%"}
P is independent of C

S is independent of C

S is independent of P, conditional on C
:::
:::::

::: aside
C = multicultural context, P = proficiency, S = code-switching.
@schachingertenes2023
:::

## Causal paths {.nostretch}

```{r}
#| label: pro-switch-1a
#| fig-asp: 0.5
#| out-width: 40%

ggplot(pro_switch_1, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(
    end_cap = ggraph::circle(15, "mm"),
    start_cap = ggraph::circle(15, "mm")
  ) +
  geom_dag_point(size = 20) +
  annotate("point", x = 2, y = 2, colour = "darkgreen", size = 20) +
  annotate("point", x = 1.5, y = 1, colour = "darkblue", size = 20) +
  geom_dag_text(size = 20 / .pt) +
  expand_plot(expand_y = c(0.1, 0.1)) +
  theme_dag(
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )
```

. . .

::: {style="text-align: center;"}
$P \rightarrow S$

$P \leftarrow C \rightarrow S$
:::

## Causal paths {.nostretch}

```{r}
#| label: pro-switch-2a
#| fig-asp: 0.5
#| out-width: 40%

ggplot(pro_switch_2, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(
    end_cap = ggraph::circle(15, "mm"),
    start_cap = ggraph::circle(15, "mm")
  ) +
  geom_dag_point(size = 20) +
  annotate("point", x = 2, y = 2, colour = "darkgreen", size = 20) +
  annotate("point", x = 1.5, y = 1, colour = "darkblue", size = 20) +
  geom_dag_text(size = 20 / .pt) +
  expand_plot(expand_y = c(0.1, 0.1)) +
  theme_dag(
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )
```

. . .

::: {style="text-align: center;"}
$S \leftarrow C \rightarrow p$
:::

## Shutting the backdoor

::: callout-tip
## Backdoor recipe

(1) List all of the paths connecting X (the potential cause of interest) and Y (the outcome).

(2) Classify each path by whether it is open or closed.
    A path is open unless it contains a collider.

(3) Classify each path by whether it is a backdoor path.
    A backdoor path has an arrow entering X.

(4) If there are any open backdoor paths, decide which variable(s) to condition on to close it (if possible).
:::

## Workplace and use of prestige variable {.nostretch}

```{r}
#| label: work_use
#| fig-asp: 0.5
#| out-width: 40%

work_use <- dagify(
  P ~ A + E + W, E ~ A + S, A ~ S, W ~ S,
  coords = tibble(
    name = c("S", "W", "E", "A", "P"),
    x = c(1, 2, 1.5, 1, 2),
    y = c(2, 2, 1.5, 1, 1)
  )
)

wu <- ggplot(work_use, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(
    end_cap = ggraph::circle(15, "mm"),
    start_cap = ggraph::circle(15, "mm")
  ) +
  geom_dag_point(size = 20) +
  annotate("point", x = 2, y = 2, colour = "darkgreen", size = 20) +
  annotate("point", x = 2, y = 1, colour = "darkblue", size = 20) +
  geom_dag_text(size = 20 / .pt) +
  expand_plot(expand_y = c(0.1, 0.1)) +
  theme_dag(
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background  = element_rect(fill = "transparent", colour = NA)
  )

wu
```

::: {.fragment fragment-index="1" style="font-size: 0.75em;"}
| Path | Open | Backdoor |
|----|:--:|:--:|
| $W \rightarrow P$ | [yes]{.fragment fragment-index="2"} | [no]{.fragment fragment-index="7"} |
| $W \leftarrow S \rightarrow A \rightarrow E \rightarrow P$ | [yes]{.fragment fragment-index="3"} | [yes]{.fragment fragment-index="8"} |
| $W \leftarrow S \rightarrow A \rightarrow P$ | [yes]{.fragment fragment-index="4"} | [yes]{.fragment fragment-index="9"} |
| $W \leftarrow S \rightarrow E \rightarrow P$ | [yes]{.fragment fragment-index="5"} | [yes]{.fragment fragment-index="10"} |
| $W \leftarrow S \rightarrow E \leftarrow A \rightarrow P$ | [no]{.fragment fragment-index="6"} | [yes]{.fragment fragment-index="11"} |
:::

::: aside
S = socio-economic status, A = age, E = exposure to prestige, W = workplace style, P = use of prestige.
:::

## Workplace and use of prestige: adjustment sets {.nostretch}

```{r}
#| label: work_usea
#| fig-asp: 0.5
#| out-width: 40%

wu
```

::: {style="text-align: center; font-size: 2em;"}
A, E

S
:::

::: {.fragment style="text-align: center; font-size: 1.5em;"}
`lm(P ~ W + S)`
:::

## References
